---
title: "Assignment 3 - Price Estimation of Diamonds"
author: "Abdullah Yıldız"
date: "9/13/2020"
output: 
  html_document:
    fig_width: 5
    fig_height: 5
    toc: true 
    toc_float:
      collapsed: no
    number_sections: true 
    theme: cerulean   # I like these theme's headers' color 
    highlight: kate
    code_folding: hide
    
---
/* Whole document: */
body{
  font-family: Helvetica;
  font-size: 16pt;
}
/* Headers */
h1,h2,h3,h4,h5,h6{
  font-size: 24pt;
}

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align = "center", message=FALSE, warning=FALSE, error = FALSE)
```

# Diamonds:

According to the [this site](http://www.diamondc.com.hk/us/factors-affect-diamond-price), there are many factors affecting the price of diamonds. Colour, Clarity, Carat (size), Cut, Shape and Fluorescence are the major determinants of the price of diamonds. Difference in single grade or level will make the price change from 5% to 30%.

Again from [this site](http://www.diamondc.com.hk/us/how-to-select-diamond), with many diamonds to select from, knowledge on diamond is necessary. However, it may be a novel experience for many people, in this section it is provided some useful information that will help to make the best decision from various choices.

To learn about diamond selection, there is one major rule to follow -- "the 4Cs'". They are Colour, Cut, Clarity and Carat. These characteristics, you can shop for a diamond within your budget.


I add many packages that may be useful to use in this analysis. But,I may not use some of them in that project.
```{r, include=FALSE}

library(dplyr)
library(ggplot2)
library(lubridate)
library(tidyverse)
library(tidyr)
library(kableExtra)
library(corrplot)
library(grid)
library(lattice)
library(ggpubr)
library(scales)
library(wesanderson)
library(RColorBrewer)
library(Metrics)
library(data.table)
library(knitr)
library(caret)
library(e1071)
library(rpart)
library(rpart.plot)
library(rattle)
library(broom)
library(cowplot)
```




# Diamonds Price Estimation

That assignment aims to find a prediction model for estimating the price of a diamond given its properties. \
`rDiamonds` data set in `rggplot2` package is used for the assignment. 

Training and test data sets are created. 

```{r Diamonds}
set.seed(503)
diamonds_test <- diamonds %>% mutate(diamond_id = row_number()) %>% 
    group_by(cut, color, clarity) %>% sample_frac(0.2) %>% ungroup()
diamonds_train <- anti_join(diamonds %>% mutate(diamond_id = row_number()), 
    diamonds_test, by = "diamond_id")
diamonds_show <- bind_rows(diamonds_train,diamonds_test)
```

Below you can see  samples from train dataand test data sets.

```{r}
head(diamonds_train)%>%
  kbl(caption = "Sample from Training Data") %>%
  kable_minimal(full_width = F)
head(diamonds_test)%>%
  kbl(caption = "Sample from Test Data") %>%
  kable_minimal(full_width = F)
  
  
```
## Data Information

In this project as you can see above table,there are some facts which can be used to do estimation. My main goal is to provide to create understandable explatory data analysis and create model for diamond price estimation.

```{r}
diamonds %>%
  glimpse



```

```{r}


sum(any(is.na(diamonds)))

```


There is `r sum(any(is.na(diamonds)))` NA values.

Since there is any  missing value in any row or column, we can get into next step :
It should be checked whether there is a duplicated row or not.



```{r}
sum(as.numeric(duplicated(diamonds)))
```

 There are `r sum(as.numeric(duplicated(diamonds)))` duplicated lines. It should be extracted these values from dataframe.

```{r}

diamonds <- unique(diamonds)

sum(as.numeric(duplicated(diamonds)))
```
 Now, we have `r nrow(diamonds)` rows and `r ncol(diamonds)` columns.Lets examine our dataset detailly such as looking data.
 


```{r}
str(diamonds)
```

As i search on the Internet, colorless diamonds has higher prices too, so that  color levels should be sorted again. 
You can fin related information in [there] (https://www.petragems.com/education/diamond-color/)
```{r}
diamonds$color <- factor(diamonds$color, levels = c("J", "I", "H", "G", "F", "E", "D"))
```

Let's look at the data again to be sure.

```{r}
str(diamonds)
```


```{r}
summary(diamonds) 
head(diamonds) 
```

##Preprocessing

Now,I start to control values of some variables such as price , dimensions of diamond (x,y,z) and etc. . All they have should have logical values as I learned from operation research in detailly. I will filter these values to make my estimation and analyze more accurate in following codes.

```{r}
diamonds %>%
  filter(price <= 0) %>%
  summarise(NegativePrice = n())
```



```{r}
diamonds %>%
  filter(x <= 0 & y > 0 & z > 0) 
diamonds %>%
  filter(y <= 0 & z > 0 & x > 0) 
diamonds %>%
  filter(z <= 0 & y > 0 & x > 0) 
```




While all `x` and `y` values are positive, Z has some negative values.


```{r}
diamonds %>%
  filter(x <= 0 & z <= 0)
```

```{r}
diamonds %>%
  filter(y <= 0 & z <= 0)
```

```{r}
diamonds %>%
  filter(x <= 0 & y <= 0 ) 
```



```{r}
diamonds = diamonds %>%
  filter(!(x == 0 & z == 0))
diamonds %>%
  filter(x == 0 | y == 0 | z == 0)
```


As It can be seen , still there are some Z values "0". After thinking how i can find z values just konwing x, y values, I thought to look at dataset again and see the superhero : "Depth". It's time to calculate Z from x,y and depth.

```{r}
diamonds = diamonds %>%
  mutate(z = ifelse(z == 0 & x != 0 & y != 0,round(depth * mean(c(x, y)) / 100, 2), z)) 
```

Controlling dataframe again

```{r}
diamonds = diamonds %>%
  filter(!(x == 0 & z == 0))
diamonds %>%
  filter(x == 0 | y == 0 | z == 0)
```

Yeah! there are any row has zero value.


Now, we can compare the x, y and z values with each other. These values should be highly correlated with each other.

```{r}
diamonds %>%
  ggplot(., aes(x = x, y = y)) +
  geom_point() + 
  geom_smooth(method = "lm")
diamonds %>%
  ggplot(., aes(x = z, y = y)) +
  geom_point() + 
  geom_smooth(method = "lm")
```

As it showed almost all these values are highly correlated except some outliers. I prefer to ignore them because I think these outliers would not affect my analysis in sight.



# EDA: Explatory Data Analysis

As I mentioned above there are 4 criterias that affects the price and quality of diamonds. Now its time to examine each of the 4Cs' :

*Colour, Cut, Clarity and Carat.*


## The  Relation Between Carat, Depth, Table, Cut with Price
- The relation between price and other variables are visualized below.
```{r,warning=FALSE,message=FALSE,class.source="fold-hide"}
plotone<-ggplot(diamonds, aes(x=carat, y=price, color=clarity)) + geom_point()+labs(x="Carat", y="Price",title="The Relation Between Carat and Price")+theme(plot.title = element_text(size=9,face="bold",vjust=5),legend.title = element_text(size = 8),legend.text = element_text(size = 7)) +
        guides(shape = guide_legend(override.aes = list(size = 1.25)),
               color = guide_legend(override.aes = list(size =3)))
plottwo<-ggplot(diamonds, aes(x=depth, y=price, color=clarity)) + geom_point()+labs(x="Depth", y="Price",title="The Relation Between Depth and Price")+theme(plot.title = element_text(size=9,face="bold",vjust=5),legend.title = element_text(size = 7),legend.text = element_text(size = 7))+
        guides(shape = guide_legend(override.aes = list(size = 1.25)),
               color = guide_legend(override.aes = list(size =3)))
plotthree<-ggplot(diamonds, aes(x=table, y=price, color=clarity)) + geom_point()+labs(x="Table", y="Price",title="The Relation Between Table and Price")+theme(plot.title = element_text(size=9,face="bold",vjust=5),legend.title = element_text(size = 7),legend.text = element_text(size = 7))+
        guides(shape = guide_legend(override.aes = list(size = 1.25)),
               color = guide_legend(override.aes = list(size =3)))
plotfour<-ggplot(diamonds, aes(x=cut, y=price, color=clarity)) + geom_point()+labs(x="Cut", y="Price",title="The Relation Between Cut and Price")+theme(plot.title = element_text(size=9,face="bold",vjust=5),legend.title = element_text(size = 7),legend.text = element_text(size = 7),axis.text=element_text(size=7))+
        guides(shape = guide_legend(override.aes = list(size = 1.25)),
               color = guide_legend(override.aes = list(size =3)))
plot_grid(plotone, plottwo,plotthree,plotfour, labels = "AUTO")
```

### Colour


```{r}

diamonds %>%
  group_by(color)%>%
  summarise(count = n()) %>%
  ggplot(., aes(x=color, y = count, fill = count)) +
    geom_col() +
    scale_fill_gradient("count", low="lightblue", high="darkblue") +
    geom_line(aes(y = count), size = 1.5, color="yellow", group = 1) +
    theme_minimal() +
    theme(legend.position = "none") +
    labs(title = "Diamonds According to the Colors",
         x = " Colour of Diamond",
         y = "Number of Colour out of all Diamond")
    
```

Bar chart show us that the most used color in diamonds are G, while the least one is J.

```{r}
diamonds %>%
  group_by(color)%>%
  summarise(color_count = n(),
            MinPrice = min(price),
            MaxPrice = max(price),
            AveragePrice = mean(price))%>%
  kable(col.names = c("Color", "Count","Minimum Price",  "Maximum Price","Average Price")) 
```

The table above illustrates the most expensive color is J by looking average prices. Also, range is maximum in G colored. According to the [this site](https://www.petragems.com/education/diamond-color/)  D-F color scale rates higher on the color grading scale than a diamond that is light yellow. Then, G-F color scale follows. So,there is a difference between theory and dataset, other facts should be take into consideration too.


### Cut



```{r}
diamonds %>%
  group_by(cut) %>%
  summarise(count = n()) %>%
  mutate(percentage = 100*count/sum(count)) %>%
  ggplot(., aes(x = '', y = count, fill = cut)) + 
    geom_bar(width = 2, stat = "identity", alpha = 1.3) +
    coord_polar("y") +
    theme_void() +
    theme(plot.title = element_text(vjust = 0.8)) +
    geom_text(aes(label = paste(format(percentage,digits=3), "%")), size=4, color = "red", position = position_stack(vjust = 0.6)) +
    labs(title = " %  of Quality of Cut ",
         fill = "Quality of the Cut")
```

The chart above illustrates that greatest majority of cut is ideal cut. When other facts are ignored, Ideals one should be more expensive.
Lets look at our max,min table again.

The surprising fact is the most expensives cuts looks like *premium and fair*




```{r}
diamonds %>%
  group_by(cut)%>%
  summarise(cut_count = n(),
            MinPrice = min(price),
            MaxPrice = max(price),
            AveragePrice = mean(price))%>%
  kable(col.names = c("Cut", "Count","Minimum Price",  "Maximum Price","Average Price")) 
```

### Clarity

Clarity is defines how much is stain existed in diamond.

```{r, fig.width=6, fig.height=4}
diamonds %>%
  mutate(clarity = factor(clarity)) %>%
  group_by(clarity) %>%
  summarise(clarity_count = n()) %>%
  
  ggplot(., aes(x=clarity, y = clarity_count, fill = clarity_count)) +
  geom_col() +
  scale_fill_gradient("clarity_count", low="peachpuff", high="peachpuff4") +
  geom_line(aes(y = clarity_count), size = 1.2, color="black", group = 1) +
  theme_minimal() +
  theme(legend.position = "none") +
  theme(axis.text.x = element_text(angle = 90), legend.position = "none")+
        labs(title = "The Number of Diamonds According to the Clarity",
         x = "Clarity",
         y = "Count")
```
As you can read from [here](https://www.bluenile.com/education/diamonds/clarity), the best clarity in that table is IF. However, average price and number of IF diamonds are very less compared to others.



```{r}
diamonds %>%
  group_by(clarity)%>%
  summarise(clarity_count = n(),
            MinPrice = min(price),
            MaxPrice = max(price),
            AveragePrice = mean(price))%>%
  kable(col.names = c("Clarity", "Count","Minimum Price",  "Maximum Price","Average Price"))
```


### Carat 

A carat is a unit of measurement used to specify the weight of a diamond. ‘Carat’ is a diamond industry special word for the weight of a diamond stone. One carat is equal to 200 milligrams, for example, a 5-carat stone will weigh 1 gram.

People often mistakenly assume that a diamond's size is synonymous with its carat weight, though that's not necessarily true. Like a human body the more the weight, the more likely the size is bigger, but other factors also affect the size. The way in which the diamond is cut will greatly affect the diameter and brilliance of the diamond. Diamonds with a poorer cut (for example, a change of depth or a thick girdle) will maintain their heavy weight but 'hidden' in the base of the diamond with less surface on top so they look smaller.


```{r}
diamonds %>%
  mutate(carat = factor(carat)) %>%
  group_by(carat) %>%
  summarise(carat_count = n())%>%
  arrange(desc(carat_count)) %>%
  head(5) %>%
  ggplot(., aes(y=carat_count, x = reorder(carat,-carat_count), fill = carat)) +
  geom_col() +
  geom_text(aes(label = carat_count), size=3, color = "black", position = position_stack(vjust = 0.60)) +
  theme_minimal() +
  scale_fill_brewer(palette = c("Paired")) +
  theme(legend.position = "none") +
  labs(title = "The Most 10 Popular Carat",
       x = "Carat", y = "Count")
```

According to the data set, the most preferable carat is 0.3 carat. In table above it seems 0.3-0.31-0.32 are some of the most existed carats.To see all of them, I also create a table you can see below. By lookig at table, it can be also seen the price effect to carats. 

```{r}
diamonds %>%
  group_by(carat) %>%
  summarise(carat_count = n(),
            MinPrice = min(price),
            MaxPrice = max(price),
            AveragePrice = mean(price))%>%
  arrange(desc(AveragePrice)) %>%
  kable(col.names = c("Carat", "Number of Carats","Minimum Price", "Maximum Price", "Average Price")) %>%
  kable_minimal (full_width = T)
```
# Create Model




```{r}
alldata <- ggplot(diamonds, aes(x=price))+
  geom_histogram  (color="aquamarine4", fill="aquamarine2")+ggtitle("Density Plot of Diamond Dataset")+theme_minimal()+ scale_y_continuous(labels = comma)
traindata <- ggplot(diamonds_train, aes(x=price))+
  geom_boxplot (color="cyan4", fill="cyan1")+ ggtitle("Density Plot of Training Set")+
  theme_minimal()+ scale_y_continuous(labels = comma)
testdata <- ggplot(diamonds_test, aes(x=price))+geom_boxplot(color="springgreen3", fill="springgreen1")+ggtitle("Density Plot of Test Set")+
  theme_minimal()+scale_y_continuous(labels = comma)
ggarrange(alldata, traindata, testdata, ncol = 1, nrow = 3)
```

It is created a main model with some variables. I choice to use backward regression which I learned in Operation Research Lesson  to choose the best  model. While starting project I already extract zero and NA values (if it existed), so I dont need to extract again.

```{r }

Main_model <- glm(price~x+y+z+carat,family = Gamma,data=diamonds_train,start=c(0.5,0.5,0.5,0.5,0.5))
summary(Main_model)
```
As you can see , significance of Z is the least. Thus, I create a new model without Z and expect an improvement in AIC score.

```{r, warning=FALSE}
Second_Model <- glm(price~x+y+carat,family = Gamma,
        data=diamonds_train,
        start=c(0.5,0.5,0.5,0.5))
summary(Second_Model)
```
Significance of Z is the least. Thus, I create third model without Y,Z . Model above shows better AIC performance by decreas,ng AIC.


In third all variables affect price significantly  with a significance level of 0.05.

```{r, warning=FALSE}
Third_Model <- glm(price~x+carat,family = Gamma,
        data=diamonds_train,
        start=c(0.5,0.5,0.5))
summary(Third_Model)
```



## Generalized Linear Model (GLM)

Here is my final model :

```{r, warning=FALSE}
Final_Model <- glm(price~x+carat,family = Gamma(link="identity"),
                  data=diamonds_train,
                  start=c(0.5,0.5,0.5))
summary(Final_Model)
```

### Residual Plot

According to residual plot, constant variance and zero mean assumptions seems to hold for the used model.
```{r, warning=FALSE }
Final_Model_Test_Data <- glm(price~x+carat,family = Gamma(link="identity"),
                  data=diamonds_test,
                  start=c(0.5,0.5,0.5))
ggplot(Final_Model_Test_Data) + 
  geom_point(aes(x=.fitted, y=.resid, col=.resid)) +
  theme_minimal() + 
  ggtitle("Residual Plot") +
  labs(x = "Fitted Values",
       y = "Residuals") +
  scale_color_gradient(low="seagreen1", high="seagreen4") +
  labs(col = "Residuals")
  
```



```{r}
Final_Model_Test_Data <- glm(price~x+carat,family = Gamma(link="identity"),
                  data=diamonds_test,
                  start=c(0.5,0.5,0.5))
ggplot(Final_Model_Test_Data) + 
  geom_smooth (aes(x=.fitted, y=.resid, col=.resid)) +
  theme_minimal() + 
  ggtitle("Residual Plot") +
  labs(x = "Fitted Values",
       y = "Residuals") +
  scale_color_gradient(low="lemonchiffon", high="lemonchiffon4") +
  labs(col = "Residuals")
  
```
  


### References

[Gamma Distribution](https://wiki.analytica.com/index.php?title=Gamma_distribution)
[Diamond Exploration Price Modeling](https://www.kaggle.com/abhishekheads/diamond-exploration-price-modeling)<br>
[EDA Example with Diamonds data set](http://rstudio-pubs-static.s3.amazonaws.com/400929_1fe468939a9c4d9c8cf8e8768ab5fb3c.html)<br>
[Wikipedia](https://en.wikipedia.org/wiki/Diamond) <br>
[Color Cheatsheet](https://www.nceas.ucsb.edu/sites/default/files/2020-04/colorPaletteCheatsheet.pdf)<br>
[Diamond Guide](https://shiningdiamonds.co.uk/diamond-buying-guide/diamond-carat)<br>
[Geom Histogram](http://www.sthda.com/english/wiki/ggplot2-histogram-plot-quick-start-guide-r-software-and-data-visualization)<br>
[Ideal Cut](https://www.diamonds.pro/education/diamond-depth-and-table/#:~:text=For%20a%20cushion%20cut%20diamond,or%2069%20to%2070%20percent)<br>

