---
title: "HW2"
output: html_document
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
 
 ##Introduction
This report is created according to the data from 1 July to 31 July  [Energy Exchange Istanbul(EXIST) page](https://rapor.epias.com.tr/rapor/xhtml/ptfSmfListeleme.xhtml). 

```{r , echo=TRUE}


setwd("C:/Users/ABDULLAH/Desktop")

library(readr)
library(tidyverse)
library(lubridate)


veri<- read.csv("HW.csv",stringsAsFactors = TRUE, header=TRUE)

str(veri)



```
So as to see changes , I use Glimpse() function. I also edited some columns in excel before starting this report. 

```{r, echo=TRUE}

veri %>% glimpse()
data <- veri %>% 
  rename(  d = Date, mcp = MCP, smp = SMP, pip = Positive.Imbalance.Price..TL.MWh., nip = Negative.Imbalance.Price..TL.MWh., sd = SMP.Direction)

data$d <- strptime(data$d,format="%d.%m.%y %H:%M") 

  

data %>% glimpse()



```
##General Information about dataset

Some statistics like min, quartile , median and max can be seen below.

Also, It can be seen that SMP values are higher than MCP values. So, we can mention about energy deficit for those.


```{r summary, message=FALSE}

data %>% summarise(count=n())
data %>% filter(sd == "Energy Deficit") %>% summarise(count=n())
data %>% filter(sd == "In Balance") %>% summarise(count=n())
data %>% filter(sd == "Energy Surplus") %>% summarise(count=n())




data %>% summary
ggplot(data, aes(x=mcp, y=smp)) + geom_point() + theme(aspect.ratio=1)+ xlim(0,350) + ylim(0,460) + ggtitle("July 2020 MCP vs SMP Prices  ") + labs(x = "mcp", y = "smp") + geom_abline(slope=1, intercept=0, color="purple")
```


###Observing MCP Hourly 

In energy industry, data is collected as hourly. In this report, data is extracted from July 1 to July 31.  The price changes regarding the hour of the day should be observed. In the graph, you can see relationship  `MCP` vs. `hour_of_day` for same hours

It can be observed that there maybe a relationship between same hours of different days,


I also create second ggplot to calculate the hourly average values for each hour.


```{r hourly, message=FALSE}
plot_one <- data %>% group_by(hour_of_day = lubridate::hour(d), day=lubridate::date(d))
ggplot(plot_one, aes(x = hour_of_day, y = mcp, color=as.character(day))) + geom_point() + ggtitle("July 2020 MCP vs Hour of Day ") + labs(x="Hour of Day", y="MCP")





plot_two<-data %>% group_by(hour_of_day = lubridate::hour(d)) %>% summarise(avg_MCP = mean(mcp), avg_SMP = mean(smp))
ggplot(plot_two) + geom_point(aes(x = hour_of_day, y = avg_MCP, color="MCP")) + geom_point(aes(x = hour_of_day, y = avg_SMP, color="SMP")) + ggtitle("Hourly Average MCP and SMP vs July 2020 Hour of Day") + labs(x="Hour of Day", y="Average MCP and SMP")



```


Here you can see average values of MCP and SMP 

```{r, message=FALSE}
plot_three <- data %>% group_by(day = lubridate::day(d)) %>% summarise(daily_avg_MCP= mean(mcp), daily_avg_SMP= mean(smp),daily_avg_pip=mean(pip),daily_avg_nip=mean(nip))
print(plot_three)
ggplot(plot_three) + geom_point(aes(x = day, y = daily_avg_MCP, color=as.character(day)))
```



The first graph shows that MCP is getting higher after morning hour by hour.


```{r , message=FALSE, warning=FALSE}
plot_four <- data %>% group_by(hour_of_day = lubridate::hour(d), day=lubridate::date(d))
ggplot(plot_four, aes(x = hour_of_day, y = day, fill=mcp)) + 
  geom_point() + 
  ggtitle("MCP Heat Map Over Hours of July 2020") +
  geom_tile() +
  scale_fill_gradient(low="white", high="red") +
  labs(x = "Hours",
       y = "Dates") +
  scale_y_date(date_breaks = "5 day", date_labels = "%m/%d") +
  scale_x_discrete(limits=c(0:23)) +
  theme(axis.text.x = element_text(angle = 45), legend.position = "top",
        legend.title = element_blank())

print(plot_four)
```
Almost all days, MCP is lower between 5-6 and 9-10 am.


```{r , message=FALSE, warning=FALSE}


plot_five <- data %>% group_by(hour_of_day = lubridate::hour(d), dayofweek = lubridate::wday(d))
ggplot(plot_five, aes(x = hour_of_day, y = dayofweek , fill=mcp)) +  ggtitle("MCP Heat Map Over Hours as Days") +

  geom_tile()
  
  
  


print(plot_five)

  

```
























